Numerical modelling of the Cordilleran ice sheet

### 2014-07-16 kappa.tex
Add corrections from Arjen.

\small{\begin{enumerate}
  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
simulations and wrote the manuscript; C.~Khroulev implemented necessary code
changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
some of the forcing data. All authors contributed to improve the text.
  \item[II] I analysed the weather data, wrote the positive degree day model,
ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
reviews.
  \item[III] I.~Rogozhina provided the original idea; I analysed the weather
data; and we wrote the manuscript together.
  \item[IV] I implemented necessary code changes in PISM, ran the simulations
and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
contributed to improve the text.
\end{enumerate}}

@@

\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}
\bigskip

As a first step towards simulations of the Cordilleran ice sheet through the
last glacial cycle, I have chosen to run the ice sheet model under a constant
climate, targeting the LGM as the period for which ice sheet limits are best
known. These simulations, presented in \CCLI, use constant temperature offsets
from the present climate state. They start from ice-free conditions and run for
10\,ka, a period chosen as representative of the ice-sheet build-up from
assumed, negligible ice cover. Although this assumption is disproved in \CCYC,
the simplistic set-up used in \CCLI\ allowed to test model sensitivity to the
choice of present-day climate forcing data, without involving too many
additional parameters.

Discrepancies between input data used (\cref{tab:datasources}) caused much
variability in modelled ice sheet geometries (\CCLI, Figs.~5--7). In
particular, continental parts of the ice-sheet are strongly dependent on input
precipitation patterns, which are often poorly resolved by atmospheric
reanalysis fields used to force the model (\CCLI, Figs.~10--14). The closest
match between model output and the reconstructed ice margin \citep{Dyke.2004}
was obtained using atmospheric forcing from the North American Regional
Reanalysis \citep[NARR,][]{Mesinger.etal.2006} and a temperature depression of
7\,\degC\ (\CCLI, Fig.~15). The relevant simulation is reproduced here
(\cref{fig:plot-lgm}), using a higher horizontal resolution of 3\,km, instead
of 10\,km as in the paper. Following this sensitivity study, I have retained
the NARR to force subsequent simulations of the Cordilleran ice sheet
presented here and in \CCYC.

@@

% ----------------------------------------------------------------------
\subsection{Daily temperature variability}
% ----------------------------------------------------------------------

\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}
\bigskip

\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}
\bigskip

Next to the atmospheric forcing, another critical source of uncertainty in
palaeo-ice sheet modelling is the choice of parameters used by the surface mass
balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
are the degree-day factors for snow, $F_s$, and ice, $F_i$, and the standard
deviation of near-surface air temperature, $\sPDD$. Although the first two can
only be deduced from laborious surface mass-balance measurements on present-day
glaciers \citep[e.g.,][]{Shea.etal.2009}, $\sPDD$ can be estimated from
temperature time-series \citep[e.g.,][]{Fausto.etal.2011}.

In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$.
The resulting grids show
that, although often assumed constant, $\sPDD$ varies greatly by location and
through the year, most particularly in the polar regions
(\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using a simple surface mass balance
model, I then demonstrated the effect of these variations on the computation of
positive degree-days (\PSDV, Fig.~2), and surface mass-balance (\PSDV, Fig.~3).

The analysis is taken further in \PSDP, using re-analysed air temperatures from
ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial resolution, the
ERA-40 time series have a longer time span than the ERA-Interim ones, which
is necessary to compute standard deviation values that are representative of
the long-term daily variability \citep{Rogozhina.Rau.2014}. In \PSDP, we
demonstrated that, over the Greenland ice sheet, there exist an inverse
correlation between $\sPDD$ and monthly mean temperatures, $T_m$
(\PSDP, Fig.~1). This correlation could potentially be used as a
parametrisation, and can be expressed as (\PSDP, Eq.~5)

@@

The effect of the PDD standard deviation, $\sPDD$, on simulations of the
Cordilleran ice sheet is not shown in any of the papers. However, following
these two studies, a natural step forward was to implement, in PISM, spatially
and seasonally variable $\sPDD$ input (PISM issue \issue{179}), and the
parametrisation expressed by \cref{eqn:sigma} over ice-filled grid cells
(\issue{265}). Their effect on the modelled Cordilleran ice sheet surface is
demonstrated here (\cref{fig:plot-sdeffect}) using a constant-climate set-up
similar to the one used in \CCLI.

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-sdmap}}
  \caption{July, daily temperature variability in the Arctic, as shown by
           monthly standard deviation of near-surface daily mean air
           temperature computed from ERA-40 reanalysis
           \citep{Uppala.etal.2005}. Note the strong differences between
           continental and maritime regions, notably within the model domain
           (black rectangle). Although daily variability is lowest in summer,
           this is when it most importantly contributes to surface melt.}
  \label{fig:plot-sdmap}
\end{figure}

@@

\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}
\bigskip

From the beginning, the primary goal of this thesis has been to perform
simulations of the Cordilleran ice sheet through the last glacial cycle. This
is the topic of \CCYC, where monthly air temperature, precipitation, and PDD
standard deviation climatologies from the NARR are amended by
palaeo-temperature time-series from six proxy records to produce a
time-dependent climate forcing (\CCYC, Fig.~2). The resulting simulations all
produce two major glaciations between MISs~4 and 2. However the differences in
timing and magnitude of these events is considerable (\CCYC, Fig.~3--4).

Geological evidence regarding the timing of glaciation allowed us to select two
simulations, driven by the GRIP \citep{Dansgaard.etal.1993} and EPICA
\citep{Jouzel.etal.2007} ice-core records, as producing more realistic
reconstructions of the last glacial cycle. I have re-run these two simulations
at a higher resolution of 6\,km, to allow for a more detailed study of
transient ice cap cover before and after the LGM, stability of major
ice-dispersal centres, and patterns of deglaciation (\CCYC, Fig.~5--15).
Modelled ice volume evolution and snapshots of ice flow patterns from one of
these two simulations are reproduced here (\cref{fig:plot-snapshots}).

@@

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-snapshots}}
  \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice
           sheet. Basal topography surface topography (1\,km contours) and
           surface velocity (m/a).
           \textbf{(b)} Evolution of total ice volume during the entire last
           glacial cycle, expressed as meters of sea-level equivalent. The
           simulation uses a horizontal resolution of 6\,km and is forced by
           atmospheric fields from the NARR \citep{Mesinger.etal.2006} and
           temperature offset time-series derived from the GRIP
           \citep{Dansgaard.etal.1993} ice-core record as described in \CCYC.}
  \label{fig:plot-snapshots}
\end{figure}

@@

In this thesis, I present numerical reconstructions of the former Cordilleran
ice sheet to a level of detail that has, to my knowledge, never been attained
before. The coupled SIA-SSA approximation is crude, yet it is today the only
approach that allows, combined with massively parallel computing, high-spatial
resolution over the multi-millenial time-scales of ice sheet growth and decay.

Through sensitivity studies, I have identified the predominant sources of
uncertainty, which I have assessed using a combination of detailed
meteorological datasets and available geological evidence. The simulations
supporting this thesis allowed estimates of the LGM Cordilleran ice sheet
extent and volume, identification of the main ice-dispersal centres, and
partial understanding of the deglaciation patterns. However, there remains much
room for model improvements and further analysis of the simulation output.

@@

Throughout the thesis, precipitation rates are prescribed as monthly, spatial
grids, but assumed equal to their present values. Although the distribution of
precipitation has been undoubtedly affected by past changes in global climate
and the intermittent presence of ice sheets and ice caps in the Northern
Hemisphere, its changes are difficult to assess. In the North American
Cordillera, much of the continental interior was likely dryer than today due to
the presence of ice caps reinforcing the topographic barrier formed by the
Pacific Coast ranges. However, some regions might have become rainier as a side
effect of redirected winds.

In meteorology, these effects are typically analysed using atmospheric
circulation models. However, I have shown in \CCLI\ that atmospheric reanalyses
which are circulation models that use data assimilation techniques, fail to
agree on the present-day pattern of precipitation in the Cordillera. The
discrepancies, due to highly mountainous terrain, have dramatic effects on the
ice-sheet-model response. Considering additional difficulties inherent to
modelling past climates, and the lack of data for assimilation, we can only
expect that palaeo-climate reconstructions will produce even higher
discrepancies in the ice-sheet-model response.

@@

Geothermal heat flux, controlling en-glacial temperature distribution and the
location of basal melt areas, was assumed constant. However, the North America
is an active subduction zone with complex geothermal patterns
\citep{Blackwell.Richards.2004}. Including this variability in the ice-sheet
model will offer new opportunities to study its output.

Most notably, modelled basal flow could be compared with glacial lineations,
one of the main sources of geomorphological evidence used to reconstruct former
ice sheets. For instance, in our simulations, large portions of the Cordilleran
ice sheet remain cold-based throughout deglaciation (\CCYC, Fig.~15). As
cold-based glaciers typically do not deform their beds, this could indicate
that some of the lineation patterns we observe in the North American Cordillera
could pre-date deglaciation. Instead, they may have been formed during an
earlier phase of advance or retreat of the ice sheet.

>>>>

\small{\begin{enumerate}
  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
simulations and wrote the manuscript; C.~Khroulev implemented necessary code
changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
some of the forcing data. All co-authors contributed to improve the text.
  \item[II] I analysed the weather data, wrote the positive degree day model,
ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
reviews.
  \item[III] I.~Rogozhina provided the original idea; I analysed the weather
data; and we wrote the manuscript together.
  \item[IV] I implemented necessary code changes in PISM, ran the simulations
and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
contributed to improve the text.
\end{enumerate}}

@@

\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}.
\bigskip

As a first step towards a simulation of the Cordilleran ice sheet through the
last glacial cycle, I have opted to run the ice sheet model under a range of
constant temperature depressions, targeting the LGM as the period for which ice
sheet limits are best known. These simulations, presented in \CCLI and run at
10\,km grid resolution, use constant temperature depressions from the present
climate state (using five different input climatologies). All model runs start
from ice free conditions and run for 10\,ka, a period chosen as representative
of Cordilleran ice sheet build-up from assumed, negligible ice cover. Although
this assumption appears unjustified from results in \CCYC, the simplistic
set-up used in \CCLI\ allowed an assessment of model sensitivity to the choice
of input climatologies, without involving too many additional parameters.

Discrepancies between the five input data sets used (\cref{tab:datasources})
caused considerable variability in modelled ice sheet geometries (\CCLI,
Figs.~5--7). In particular, ice growth in continental sectors of the modelling
domain is strongly dependent on input precipitation patterns, which are often
poorly resolved by atmospheric reanalysis fields used to force the model
(\CCLI, Figs.~10--14). The closest match between modelled ice sheet extent and
the reconstructed LGM ice margin \citep{Dyke.2004} was obtained using the North
American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006} climatologies
and a temperature depression of 7\,K (\CCLI, Fig.~15). The relevant simulation
is reproduced here, using a higher horizontal resolution of 3\,km
(\cref{fig:plot-lgm}). Following this sensitivity study, I have retained the
NARR climatology as input to simulations of the Cordilleran ice sheet presented
here and in \CCYC.

@@

% ----------------------------------------------------------------------
\subsection{Importance of daily temperature variability for ice sheet modelling}
% ----------------------------------------------------------------------

\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}.
\bigskip

\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}.
\bigskip

Next to the atmospheric forcing, another critical source of uncertainty in
palaeo-ice sheet modelling is the choice of parameters used by the surface mass
balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
are the degree-day factors for snow, $F_s$, and ice, $F_i$
(\cref{fig:model-pdd, tab:surfaceparams}), and the standard deviation of
near-surface air temperature, $\sPDD$ (\cref{eq:calovgreve,
tab:surfaceparams}). Although the first two can only be deduced from surface
mass-balance measurements on present-day glaciers
\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
series \citep[e.g.,][]{Fausto.etal.2011}.

In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$. The
resulting grids show that, although often assumed constant in ice sheet
modelling, $\sPDD$ exhibits significant spacial and temporal variation, most
particularly in the polar regions (\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using
a simplified surface mass balance model, I then demonstrated the effect of this
variation on the computation of positive degree-days (\PSDV, Fig.~2), and
surface mass-balance (\PSDV, Fig.~3). Neglecting this variability in ice sheet
modelling introduces significant errors in modelled PDD and surface mass
balance calculations. The transient climate simulations presented in \CCYC\
therefore now include this effect.

The analysis is taken further extended in \PSDP, using re-analysed air
temperatures from ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial
resolution than the ERA-Interim reanalysis used in \PSDV, the ERA-40 time
series span over longer time, which is necessary to compute standard deviation
values that are representative of the long-term daily variability
\citep{Rogozhina.Rau.2014}. In \PSDP, we demonstrated an inverse correlation
between $\sPDD$ and monthly mean temperatures, $T_m$, over the Greenland ice
sheet (\PSDP, Fig.~1). This correlation could potentially be used as a
parametrisation, and can be expressed as (\PSDP, Eq.~3)

@@

The effect of the PDD standard deviation, $\sPDD$, on simulations of the
Cordilleran ice sheet is not shown in any of the papers. However, following
these two studies, a natural step forward was to implement, in PISM, spatially
and seasonally variable $\sPDD$ input (PISM issue~\issue{179}), and the
parameterisation expressed by \cref{eqn:sigma} over ice-covered grid cells
(PISM issue~\issue{265}). Their effect on modelled Cordilleran ice sheet
surface elevations is demonstrated here using a constant climate set-up similar
to the one used in \CCLI\ (\cref{fig:plot-sdeffect}).

\todo{You need to insert a couple of sentences of what you learn from Figure
      16!}

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-sdmap}}
  \caption{July, daily temperature variability in the Arctic, as shown by
           monthly standard deviation of near-surface daily mean air
           temperature computed from ERA-40 reanalysis
           \citep{Uppala.etal.2005}. Note the strong differences between
           continental and maritime regions, notably within the modelling domain
           (black rectangle). Although daily variability is lowest in summer,
           this is when it most importantly contributes to surface melt and,
           therefore, net mass balance.}
  \label{fig:plot-sdmap}
\end{figure}

@@

\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}.
\bigskip

The primary goal of this thesis has been to perform simulations of the
Cordilleran ice sheet through the last glacial cycle. This is the topic of
\CCYC, where monthly air temperature, precipitation, and PDD standard deviation
climatologies from the NARR are combined with palaeo-temperature time-series
from six proxy records to produce a time-dependent climate forcing (\CCYC,
Fig.~2). The resulting simulations, using horizontal resolutions of 6 and
10\,km, all produced two major glaciations during the last glacial cycle, at
MIS~4 and MIS~2. However, the differences in timing and magnitude of these
glaciations are considerable (\CCYC, Figs.~3--4).

Geological evidence for the timing and extent of maximum glaciation durng the
LGM allowed me to select two simulations, driven by the GRIP
\citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice core
records, as producing the most realistic reconstructions of the last glacial
cycle. I re-ran these two simulations at a higher resolution of 6\,km, in order
to allow for a more detailed study of transient ice caps before and after the
LGM, of the stability of major ice-dispersal centres, and of patterns and rates
of deglaciation (\CCYC, Figs.~5--15). Modelled ice volume evolution (expressed
as meters of sea level equivalent) and snapshots of ice extent and ice flow
patterns for the GRIP simulation (\cref{fig:plot-snapshots}) shows...
\todo{complete}.

@@

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-snapshots}}
  \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice
           sheet. Basal topography, surface topography (1\,km contours) and
           surface velocity (m/a) are model results.
           \textbf{(b)} Evolution of total ice volume during the last glacial
           cycle, expressed as metres of sea-level equivalent (m~s.l.e.). The
           simulation is performed on a 6\,km horizontal grid and adopts
           atmospheric fields from the NARR \citep{Mesinger.etal.2006} and
           temperature offset time-series derived from the GRIP ice-core record
           \citep{Dansgaard.etal.1993} as described in \CCYC. Note that, for
           this run, the Cordilleran ice sheet was almost as voluminous during
           MIS~4 as it was during the LGM.}
  \label{fig:plot-snapshots}
\end{figure}

@@

In this thesis, I present numerical reconstructions of the former Cordilleran
ice sheet to a level of detail that has, to my knowledge, not been attained
before. The coupled SIA-SSA approximation is crude, yet it is currently the
only approach that allows, combined with massively parallel computing, high
spatial resolution over the multi-millenial time scales of ice sheet growth and
decay.

Through sensitivity studies, I have identified the predominant sources of
uncertainty, which I have assessed using a combination of detailed
meteorological datasets and available geological evidence.\todo{insert a
statement which these sources of uncertainties are} Simulations supporting this
thesis yielded estimates of the LGM Cordilleran ice sheet extent and volume,
identification of the main ice-dispersal centres througout the last glacial
cycle, and understanding of final deglaciation patterns. However, there remains
considerable room for model improvements and further analysis of the simulation
output \todo{introduce your perspectives here}.

@@

One of the predominant uncertainties, for which Cordilleran ice sheet modelling
is highly sensitive, is the treatment of precipitation. Throughout the
modelling, we base precipitation patterns on current patterns derived from
reanalysis data. Although the distribution of precipitation has undoubtedly
been affected by past changes in global climate and the intermittent presence
of ice sheets and ice caps in the Northern Hemisphere, its changes are
difficult to assess. In the North American Cordillera, much of the continental
interior was likely dryer than today due to the presence of ice caps on the
Pacific Coast ranges, reinforcing the current topographic barrier. However,
some regions were probably rainier as a side effect of modified wind patterns.

In meteorology, these effects are typically analysed using atmospheric
circulation models. However, I have shown in \CCLI\ that different atmospheric
reanalyses which are circulation models that use data assimilation techniques,
fail to agree regarding the present-day pattern of precipitation over the
modelling domain. These discrepancies, which arise as a consequence of the
highly mountainous terrain, have dramatic effects on the ice sheet model
response. Considering additional difficulties inherent to modelling past
climates, and the lack of data for assimilation, we must regard derived ice
sheet histories as a highly generalized model output [only expect that
palaeo-climate reconstructions will produce even higher discrepancies in the
ice-sheet-model response].

\todo{think tomorrow. can one say something about the robustness of some
      features (ice caps that form, spreading centers which persist)which occur
      irrespective of which reanalysis is used?}.

@@

Geothermal heat flux, contributing to the en-glacial temperature distribution
and the location of basal melt areas, was assumed constant over the modelling
grid. However, the geology of the study area consists of an amalgamation of
former subduction zone, resulting in complex geothermal patterns
\citep{Blackwell.Richards.2004}. I am now at the stage to start including this
variability as a driver to the ice sheet model, which will offer new
opportunities to study its output in terms of ice flow patterns, consistent
basal freezing and basal melting patterns, regions prone for glacial erosion,
and regions characterised by glacial deposition.

I expect that inclusion of geothermal heat patterns allow modelled basal flow
directions and distances to be compared with glacial lineations, which is one
of the main sources of geomorphological evidence used to reconstruct former ice
sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
instance, in our simulations, large portions of the Cordilleran ice sheet
remain cold-based throughout deglaciation (\CCYC, Fig.~15). As cold-based
glaciers typically do not impact their beds significantly, this could indicate
that some of the lineation patterns we observe within these zones of consistent
cold-based deglaciation could be older features, formed during an earlier phase
of advance or retreat of the ice sheet. However, without a better control on
the geothermal heat flux, such analysis are perhaps too premature.
